---
output:
  html_document:
    keep_md: yes
---
## Peer assessment 2 - Exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database

### Download and preprocessing data

Load needed packages
```{r, message = F}
library(R.utils)
library(ggplot2)
```


Set work directory, download and extract bzipped dataset
```{r}
setwd("~/NOAAstorm_PeerAssesment2")
dir <- dir()
fileex <- dir=="NOAAstorm.csv.bz2"
fileex1 <- sum(fileex*1)
if (fileex1==0) {
        url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
        download.file(url,"NOAAstorm.csv.bz2",method="curl")        
        }
fileex2 <- dir=="NOAAstorm.csv"
fileex3 <- sum(fileex2*1)
        if (fileex3==0) {
        bunzip2("NOAAstorm.csv.bz2","NOAAstorm.csv",remove=FALSE)
        } 
db <- read.csv("NOAAstorm.csv", na.strings = "")
```
Download Storm Data Documentation
```{r}
if (!file.exists("Storm.Data.Documentation.pdf")) {
    url1 = "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01    016005curr.pdf"
    download.file(url1,"Storm.Data.Documentation.pdf",method="curl")
}
```


### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Select variables to answer the question
```{r}
db1 <- db[,c("BGN_DATE","STATE","EVTYPE","FATALITIES","INJURIES")]
summary (db1)
```

Create a variable that sum fatalities and injuries and eliminate events without harm
```{r}
db1$harm <- db1$FATALITIES + db1$INJURIES
summary(db1$harm)
db2 <- db1[db1$harm > 0,]
```

Summarize harm for each event type
```{r}
eventharm.df <- aggregate(harm ~ EVTYPE, data = db2, FUN = "sum")
```

Extract the top 20 causes of harm, aggregate thunderstorm and current rip data, and make a pie graph 
```{r, fig.height = 8,fig.width = 12}
top <- eventharm.df[order(-eventharm.df$harm),]
top$harm1 <- top$harm
top$harm1[3] <- top$harm[3] + top$harm[9] + top$harm[16]
top$harm1[19] <- top$harm[19] + top$harm[21]
top <- top[order(top$harm1,decreasing=T),][c(-9,-17,-21),]
top[2,1] <- "THUNDERSTORM WIND" 
top20 <- top[1:20,]
top20reordered <- top20[order(top20$harm1),]
par(mai=c(1.5,2.1,1,1))
barplot(top20reordered$harm1, horiz = T, names.arg = top20reordered$EVTYPE, las = 1,xlim=c(0,100000), cex.names = 1, xlab = "Total harmful events", main = "Top 20 causes and number of harmful events (fatalities + injuries) \nfrom 1950 to 2011 in U.S.", sub = "Data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database")
```


```{r}
db$date <- strptime(as.character(db$BGN_DATE),"%m/%d/%Y %H:%M:%S")
```

